The effect of symbols in government communications on citizens' trust, and its interaction with substantive information

### 2018-06-12 paper2_information and symbols.Rmd
Add files via upload

---
title: 'The effects of symbols and information in government communications on citizens  trust: Evidence from a survey experiment'
author: '**Saar Alon-Barkat**, PhD candidate; the Hebrew University of Jerusalem;   saar.barkat@mail.huji.ac.il'
date: " "
output:
  html_document:
    theme: flatly
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  pdf_document:
    toc: yes
  word_document: default
link-citations: yes
bibliography: phd_paper_2.bib
urlcolor: blue
---

@@

Paper presented at the 2018 Transatlantic Dialogue (TAD 14) at Workshop 1: *Behavioral Public Administration and Psychological perspectives*. June 2018, Bucharest, Romania.

@@

*Effective public communications have been proposed as a remedy for citizens' distrust in government. The studies that explored the effectiveness of government communications have focused mainly on the extent to which they provide substantive information about bureaucracies' performance and policies. Recently, a few studies pointed to the effect of symbolic elements, entangled in these communications (e.g. logos, images and celebrities). Still, we lack theoretical understanding of the different mechanisms underlying citizens' responses to information and symbolic elements, and accordingly - the interaction between them. Building on the Elaboration Likelihood Model (ELM) from social psychology, I hypothesize that symbolic elements are likely to lead citizens to dedicate less attention to the substantive information in the communication, and accordingly, to increase their trust in logically-unpersuasive policy plans. I also hypothesize that symbolic elements have a greater effect on citizens' trust the less citizens perceive the issue as personally relevant. To test these hypotheses, I conducted a survey experiment, focusing on communications of the Israeli Environment Protection Ministry, about its policy plans. The results confirm that symbolic elements in communications increase citizens' trust in ill-perceived policies, by distracting them from scrutinizing their content. In contrast with my expectation, the effect of symbols was not attenuated by personal relevance. I discuss the implications of these findings for democratic responsiveness and accountability.*

@@

The public administration research on the effect of government public communications on citizens attitudes has focused mostly on the *informative* aspects of these communications, whereby government organizations provide citizens with evidence about their performance and details and explanations about their policies and actions [e.g. @marvel_2015; @im_2014; @porumbescu_2017; @porumbescu_2016; @porumbescu_2017a; @porumbescu_2017b; @grimmelikhuijsen_2015; @ho_2017]. Public communications are expected to increase citizens' trust insofar as they provide citizens with favorable information on organizations' performance and logically persuasive explanations for their policies. At the same time, experiment-based studies that explored the responses of citizens to substantive information revealed cognitive biases that distort their processing of information in communications, and undermine its effectiveness [@baekgaard_2016; @grimmelikhuijsen_2013; @grimmelikhuijsen_2014; @hvidman_2016; @james_2011; @james_2014; @james_2017; @marvel_2015; @olsen_2017; @james_2017a; @grimmelikhuijsen_2017].

@@

This research gap has normative significance. Whereas the responsiveness of citizens to substantive information is perceived favorably from a normative view of democratic theory, the potential effect of symbolic elements can be problematic. One can argue that shaping citizens' perceptions of governments by evoking associations and emotions stands in tension with the normative expectation that citizens' judgments should be based on their critical thinking. Mainly, we should be concerned about the possibility that symbolic communication may elicit undue trust in poorly performing organizations and in ill-conceived policies, and enable public organizations to escape justified public criticism [@alonbarkat_2017]. Therefore, this study focuses on the following questions: *What are the psychological mechanisms by which symbolic elements in governemnt communications affect citizens' trust in government organizations and their policies?* Accordingly, *under what conditions do symbolic elements  increase citizens' trust in governemnt policy?* And specifically, *does the effect of symbolic elements depend on the substantive content of communications?* *Do symbols offset the rational, negative effect of unfavorable or logically unpersuasive information?*

To address these research questions, I draw upon the social psychology Elaboration Likelihood Model (ELM). ELM provides a comprehensive theoretical framework for the varying psychological processes underlying individuals' responses to persuasive communication, and their consequences for attitude change  [@petty_1986; @petty_2011]. Based on ELM, I suggest that symbolic elements in public communications can affect citizens' trust via two mechanisms: *First*, they may evoke positive feelings and associations that are transferred from the symbols to the relevant government entities and policies, and thereby increase citizens' trust in them. I expect that this effect is attenuated the more citizens perceive the communications as personally relevant, because they are more motivated to think about their messages. *Second*, symbolic elements may increase citizens' trust in government entities and policies by causing them to disregard negative of logically unpersuasive content that would have otherwise decreased their trust.

To empirically test these theoretical expectations, I conducted a large randomized survey experiment among circa `r SVIVA2_01%>%nrow() %>% round(-1)` Israeli citizens. The experiment focuses on the empirical case of the Israeli Environment Protection Ministry. I explore citizens' responses to public communications about two policy plans of the Ministry, and specifically the extent to which their trust in these policies are affected by the substantive information of the communications (logically persuasive vs. unpersuasive policy plans) and by symbolic elements in them (real symbols vs. fake symbols and no symbols conditions). In addition, I examine the moderating role of the perceived personal relevance of these policies.

I find that the symbolic elements increased citizens' trust in these policies. In line with my theoretical expectations, the effect was greater among those who saw the logically unpersuasive policy plans, who were distracted by the symbols and thus were less inclined to think about its content. I did not find support for the hypothesis that the effect of the symbolic elements is attenuated by the perceived personal relevance of the policy issue. I discuss the theoretical and normative implications of these findings.

@@

@alonbarkat_2017 examined Israeli citizens' trust in the Israel Electricity Corporation, a state owned company, and it's shaping by their exposure to comic cartoon figures that were routinely presented in the organization's public campaigns. They compared between citizens who were personally exposed to prolonged power outages, and others who were not. They found that the symbolic element had a positive effect on citizens' trust in the organization, including among those who personally experienced poor performance (a modest effect size equivalent to 0.25 standard deviations). Building on social psychology theory and marketing research, they suggested that the symbols enhance trust by inducing an unconscious transfer of affect from the symbols to the organizations and policies associated with them (evaluative conditioning). They proposed that the effectiveness of government symbolic elements depends on their familiarity, and aesthetic design, as well as on the ambiguity of the assessed government quality.

@@

Even though these studies substantiate a link between the appearance of symbolic elements in communications and citizens' favorable attitudes, they did not directly investigate the causal mechanisms that underlie this effect. Therefore, we still lack understanding of the circumstances under which symbols are likely to be more/less effective in shaping citizens' trust. Additionally, these studies neither theorized nor empirically tested the interaction between symbolic elements and citizens’ processing of substantive information in communications. We do not know whether well-designed symbols reinforce, or offset, citizens' rational responses to information. Specifically, as explained above, it is essential to assess the possibility that symbols can unduly compensate for unfavorable or logically unpersuasive information that should yield negative responses.

@@

According to ELM, peoples' responses to different elements in persuasive communications depends on the amount of thinking that they dedicate to the message content. Persuasion can occur when thinking is "high" or "low", but the processes and consequences are likely to differ in each situation. This range between high and low degrees of thought is referred to as the "elaboration continuum". Along this continuum, there are numerous specific processes of persuasion (or attitude change) that can occur. At the high end of the elaboration continuum, also called the "central route", people are likely to respond by cognitively scrutinizing the arguments central to the merits of the issue. At the low end, also called "the peripheral route" people respond to simple cues automatically and unconsciously through primitive psychological processes that require very little amount of cognitive effort, such as a transfer of affect[^footnote_dual_process_theories]. These different persuasion mechanisms are likely to cause divergent types of attitudinal change. Persuasion through the central route is expected to be more enduring, resistant and predictive of behavior, compared with peripheral route persuasion. Of course, much of the time, people are located in between the two ends of the continuum, and thus persuasion is determined by a mixture of these processes.

@@

People at different locations along the elaboration continuum may process the same element in a communication in different ways. Consider for instance a text that contains a series of logical arguments supporting a certain position. Those at the high end of the continuum would regard this text as arguments, and systematically scrutinize it based on its logical persuasive quality (central route processing). Conversely, those at the low end of the continuum are likely to disregard the content of the text, and instead, rely on the number of lines in the text, for example, as a cue for its strength (peripheral route processing). Some elements provide potential cues that lack any logical quality at all, such as the physical attractiveness of the source. These latter elements may affect the attitudes of those at the low end, but are less likely to persuade those at the high end [@petty_1986].

ELM further postulates that the placement of people along the elaboration continuum in any given situation is determined by their motivation and ability to think about the message. The literature proposes a variety of such motivational and ability factors. Most prominently, the more people perceive the issue as personally relevant, the more motivated they are to invest in thinking about it, and accordingly, they more likely they are to use the central route [@petty_1979; @petty_1981; @petty_1983]. Also, studies have demonstrated that the more people are distracted, the more difficult it is  for them to scrutinize the message arguments, a mechanism that has been referred to as "elaboration disruption" [@petty_1976; @petty_1986]. People's location along the elaboration continuum is also determined by their affective state or mood. Positive mood (compared with neutral or negative moods) reduces people's cognitive ability and attenuates their motivation to think, due to their desire to maintain their positive mood and avoid message processing that would depress it [@worth_1987; @mackie_1989; @bless_1990; @schwarz_1991; @petty_2015; but see: @wegener_1995].[^footnote_elm_factors]

@@

People's motivation and/or ability to think is shaped not only by factors external to the communication, but also by elements in the communication itself. For example, using second-person pronouns in the message (i.e. "you", "your") increases people's perceived personal relevance, and motivation to elaborate [@burnkrant_1989]; Background music in commercials can distract people and prevent them from scrutinizing their content [@park_1986]; The use of humor in television shows stimulates positive mood and reduces people's ability and motivation to scrutinize their substantive arguments [@nabi_2007; @young_2008]. Affecting the amount of thinking can cause attitude change in interaction with elements in the communication that can function as arguments or peripheral cues. Accordingly, decreasing the amount of thinking enhances the effect of peripheral cues, whereas increasing the amount of thinking enhances the effect of the logical arguments.

Finally, some elements in communications can function both as variables that affect the amount of thinking and as potential peripheral cues or arguments. Thus, these elements can shape people's attitudes through more than one causal mechanism. Continuing with the above mentioned examples, background music and humor can function at the same time as factors that reduce elaboration and as peripheral cues. @petty_1986 have suggested that we are more likely to find these complex relations of multiple mechanisms in those situations where people are located in the middle of the elaboration continuum. When people are neither extremely motivated or unmotivated to think, they may rely on different elements in the communications to decide if the message is worth or needs considering.

@@

In light of the conceptualization of ELM, substantive information elements in government public communications represent issue-related arguments, which can be scrutinized or processed systematically based on their logical quality (via the central route), at least by those who are sufficiently motivated to do so. Conversely, symbolic elements represent potential peripheral cues, which can affect attitudes through the peripheral route, by evoking positive emotions and associations that are attributed to government organizations and their policies [@alonbarkat_2017]. Accordingly, the effect of symbolic elements on citizens' trust is likely to depend on citizens' motivation to cognitively invest and think about the policy issues. Specifically, I expect that the more citizens perceive the policy issue as personally relevant, the more they are likely to focus on scrutinizing the content of the message, and the less they are likely to rely on the symbolic elements entangled in the communication.

Accordingly, I expect that:

@@

>*H<sub>2</sub> - The positive effect of symbolic elements on citizens' trust is likely to be attenuated the more citizens perceive the policy issues in the communications as personally relevant.*

However, we can also derive from ELM a second mechanism by which symbolic elements may increase citizens' trust. In addition to their direct positive effect on trust (via the peripheral route), symbols can decrease the amount of thought citizens dedicate to the communication, and thus prevent them from scrutinizing negative information (e.g. information on poor performance) and weak arguments (e.g. logically unpersuasive description of their policy plans).

Symbolic elements can decrease citizens' ability and motivation to think about the message in several ways. First, they can cause elaboration disruption, by distracting citizens and absorbing their attention and cognitive efforts [@petty_1976; @petty_1986]. Second, they can decrease citizens' motivation and ability to think about the message content by creating a positive affect and placing citizens in a positive mood. As explained above, the postulation that positive mood decreases elaboration has been explained, in part, by a mood maintenance mechanism [@mackie_1989; @bless_1990; @schwarz_1991]. This mechanism is particularly relevant to the context of government communications, in which thinking about the content of the message - government organizations and policies - is likely to yield negative mood among most people [@marvel_2015; @marvel_2015a; @hvidman_2016]. Therefore, when citizens are exposed to symbolic elements that evoke positive feelings, they may seek to avoid thinking about tedious bureaucratic affairs, which would depress their positive mood.

The expectation that symbols will reduce the amount of thinking entails an interaction between symbols and the substantive information in the communications. Symbolic elements are likely to increase citizens' trust insofar as they reduce citizens' inclination to scrutinize information that would have decreased their trust. Accordingly, I expect that:

>*H<sub>3</sub> - The positive effect of symbolic elements on citizens' trust is likely to be enhanced when the communication includes negative or logically unpersuasive information that would have decreased citizens' trust.*

@@

Analytically, the expectation that symbolic elements decrease message elaboration also entails that symbols may decrease citizens' trust in communications about highly positive performance and persuasive policy plans. On the other hand, one can make the argument that neutralizing the effect of "positive" information with symbols may not actually yield significant differences, for two reasons: First, citizens are more likely to "punish" bureaucracies for publishing negative information than to "praise" them for publishing positive information [@james_2011; @nielsen_2016; @james_2017b]. Congruently, the effect of positive information (versus a medium level) is likely to be relatively small, and hence neutralizing this effect may not make much difference. Second, even when communications include favorable information, symbols can still have a positive effect on trust through the peripheral route (in accordance with my first hypothesis). Given the focus of this study on the interaction between symbols and ill-conceived policies, the research design does not directly addresses this latter interaction, which can be .

@@

@@

Empirically, I focused on the case of the Israeli Environment Protection Ministry (hereafter: EPM), a national level bureau responsible for the formulation, co-ordination and execution of environmental policy in Israel, both at the national and local levels. I analyzed the effects of informative and symbolic elements in EPM's communications of two policy plans on citizens' trust in these policies, and the extent to which they elaborated on them. In the survey, the communications about the two policy plans, in which I manipulated the informative and symbolic elements, are presented to the participants as sections from the EPM's official publication of its annual work plan. Methodologically, EPM makes a good case study for two main reasons.

*First*, EPM has invested greatly over the past years in public communications. In these various communications, the EPM used an array of symbolic elements, which became widely recognized by the public. Here, I decided to focus on three prominent symbolic elements used in the EPM's public communications, and displayed below in **Figure 2**. The first symbolic element is the unique brand logo of the EPM (image a. in figure 1). This logo consists of pair of green and orange leaves that resemble two hands, designed to symbolize peace and harmony with the environment. The second symbolic element in which I focused is the green color, which is strongly associated with taking care of the environment. The EPM's logo and the green color can be seen in almost all of the visual communications of the EPM, and they represent key elements of its unique "visual identity" [@vandenbosch_2005]. The third element is celebrity endorsements. Specifically, I selected two high-profile celebrity comedians, *Tal Friedman* and *Ido Rosenblum*, that previously endorsed two salient advertising campaigns of the EPM. The campaign starring the comedian Tal Friedman (image b.) was launched in 2010, and included a series of advertisements in which Friedman was presented with a green grass head, and accordingly the campaign's slogan was: "starting to think green". The campaign starring the comedian Ido Rosenblum (image c.) was launched in 2017, and was focused on reducing the use of disposable bags.[^footnote_campaigns] The two campaigns were highly successful in terms of their recognizability and likability, as suggested by their evaluation reports by the Government advertising agency.

@@

The *second* reason for selecting EPM regards its policy plan regarding the pollution in the Haifa-Bay area, which offers a fortunate case for examining the moderating role of perceived personal relevance. The Haifa-Bay area, located on the Mediterranean coast of northern Israel, is characterized by relatively high levels of air pollution, mainly due to the concentration of heavy industry. Beginning in the late 2000s, the Ministry of Environment protection has been implementing policy programs to reduce the high levels of pollution in the area. The uniqueness of this policy issue is rooted in the fact that it represents natural variation in citizens' perceived personal relevance (between those who reside in the polluted area, and others), which is not strongly confounded with differences in personal knowledge, as in most cases [@petty_1983]. The air pollution in the Haifa-bay area  is an intangible, invisible, ambiguous and complicated policy issue. Those who reside in a polluted area, are not likely to have much greater knowledge and understanding of the issue. Accordingly, when they are exposed to a communication about this personally relevant issue, they are more motivated to scrutinize it, yet are not more capable to do so than others. Hence, this case enables to isolate the distinct effect of personal relevance. As explained below, the sample population of the experiment consists of residents of Haifa-bay area, matched with residents of cities at the center of Israel with similar aggregate socio-economic profile.

I supplemented the Haifa-bay air-pollution policy with a second policy plan, which is likely to be perceived as equally relevant for those Haifa-bay residents and others. This comparison further enables me to confirm that differences between the two groups, if found, are explained by differences in the perceived personal relevance of the policy issue. Furthermore, testing the effects of symbolic elements and information on two different policy plans provides a more rigorous test for the theory. Hence, I selected EPM's policy for reducing of domestic waste and increasing recycling, a tangible policy issue which I expect to be perceived as equally personally relevant by both groups of citizens. For the sake of simplicity, I henceforth refer to these two policy plans as "air-pollution" and "recycling" policies.

>>>>

---
title: "The effect of symbols in government communications on citizens' trust, and its interaction with substantive information"
author: '**Saar Alon-Barkat**, PhD candidate; the Hebrew University of Jerusalem;   saar.barkat@mail.huji.ac.il'
date: " "
output:
  html_document:
    theme: flatly
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  pdf_document:
    toc: yes
  word_document: default
link-citations: yes
bibliography: phd_paper_2.bib
urlcolor: blue
---

@@

@@

*Effective public communications have been proposed as a remedy for citizens' distrust in government. The studies that explored the effectiveness of government communications have focused mainly on the extent to which they provide substantive information about bureaucracies' performance and policies. Recently, a few studies pointed to the effect of symbolic elements, entangled in these communications (e.g. logos, images and celebrities). Still, we lack theoretical understanding of the mechanisms underlying citizens' responses to symbolic elements, and the interaction between symbols and substantive information. Building on the Elaboration Likelihood Model (ELM) from social psychology, I hypothesize that symbolic elements can increase citizens' trust via two mechanisms: First, directly, by evoking positive feelings that are transferred to the organizations and policies presented in the communication. Second, indirectly, by leading citizens to dedicate less attention to the substantive information in the communication, and thus causing them to disregard negative of logically unpersuasive content that would have otherwise decreased their trust. I examine these hypotheses via a survey experiment, focusing on communications of the Israeli Environment Protection Ministry about its policy plans. The results confirm that symbolic elements in communications increase citizens' trust. The effect was greater among those who saw the logically unpersuasive policy plans, who were distracted by the symbols and thus were less inclined to scrutinize their content. I discuss the implications of these findings for democratic responsiveness and accountability.*

@@

The public administration research on the effect of government public communications on citizens attitudes has focused mostly on the *informative* aspects of these communications, whereby government organizations provide citizens with evidence about their performance and details and explanations about their policies and actions [e.g. @marvel_2015; @im_2014; @porumbescu_2017; @porumbescu_2016; @porumbescu_2017a; @porumbescu_2017b; @porumbescu_2018; @grimmelikhuijsen_2015; @ho_2017]. Public communications are expected to increase citizens' trust insofar as they provide citizens with favorable information on organizations' performance and logically persuasive explanations for their policies. At the same time, experiment-based studies that explored the responses of citizens to substantive information revealed cognitive biases that distort their processing of information in communications, and undermine its effectiveness [@baekgaard_2016; @grimmelikhuijsen_2013; @grimmelikhuijsen_2014; @hvidman_2016; @james_2011; @james_2014; @james_2017; @marvel_2015; @olsen_2017; @james_2017a; @grimmelikhuijsen_2017].

@@

This research gap has normative significance. Whereas the responsiveness of citizens to substantive information is perceived favorably from a normative view of democratic theory, the potential effect of symbolic elements can be problematic. One can argue that shaping citizens' perceptions of governments by evoking associations and emotions stands in tension with the normative expectation that citizens' judgments should be based on their critical thinking. Mainly, we should be concerned about the possibility that symbolic communication may elicit undue trust in poorly performing organizations and in ill-conceived policies, and enable public organizations to escape justified public criticism [@alonbarkat_2017]. Therefore, this study focuses on the following questions: *What are the psychological mechanisms by which symbolic elements in governemnt communications affect citizens' trust in government organizations and their policies?* Accordingly, *does the effect of symbolic elements depend on the substantive content of communications?* *Do symbols offset the rational, negative effect of unfavorable or logically unpersuasive information?*

To address these research questions, I draw upon the social psychology Elaboration Likelihood Model (ELM). ELM provides a comprehensive theoretical framework for the varying psychological processes underlying individuals' responses to persuasive communication, and their consequences for attitude change  [@petty_1986; @petty_2011]. Based on ELM, I suggest that symbolic elements in public communications can affect citizens' trust via two mechanisms: *First*, they may evoke positive feelings and associations that are transferred from the symbols to the relevant government entities and policies, and thereby increase citizens' trust in them. *Second*, symbolic elements may increase citizens' trust in government entities and policies by causing them to disregard negative of logically unpersuasive content that would have otherwise decreased their trust.

To empirically test these theoretical expectations, I conducted a large randomized survey experiment among circa `r SVIVA2_01%>%nrow() %>% round(-1)` Israeli citizens. The experiment focuses on the empirical case of the Israeli Environment Protection Ministry. I explore citizens' responses to public communications about two policy plans of the Ministry, and specifically the extent to which their trust in these policies are affected by the substantive information of the communications (logically persuasive vs. unpersuasive policy plans) and by symbolic elements in them (real symbols vs. fake symbols and no symbols conditions).

I find that the symbolic elements increased citizens' trust in these policies. In line with my theoretical expectations, the effect was greater among those who saw the logically unpersuasive policy plans, who were distracted by the symbols and thus were less inclined to think about its content. I discuss the theoretical and normative implications of these findings.

@@

@alonbarkat_2017 examined Israeli citizens' trust in the Israel Electricity Corporation, a state owned company, and it's shaping by their exposure to comic cartoon figures that were routinely presented in the organization's public campaigns. They compared between citizens who were personally exposed to prolonged power outages, and others who were not. They found that the symbolic element had a positive effect on citizens' trust in the organization, including among those who personally experienced poor performance (a modest effect size equivalent to 0.25 standard deviations). Building on social psychology theory and marketing research, they suggested that the symbols enhance trust by inducing an unconscious transfer of affect from the symbols to the organizations and policies associated with them ("evaluative conditioning"). They proposed that the effectiveness of government symbolic elements depends on their familiarity, and aesthetic design, as well as on the ambiguity of the assessed government quality.

@@

Even though these studies substantiate a link between the appearance of symbolic elements in communications and citizens' favorable attitudes, they did not directly investigate the causal mechanisms that underlie this effect. Therefore, we still lack understanding of the circumstances under which symbols are likely to be more/less effective in shaping citizens' trust. Specifically, these studies neither theorized nor empirically tested the interaction between symbolic elements and citizens’ processing of substantive information in communications. We do not know whether well-designed symbols reinforce, or offset, citizens' rational responses to information. Specifically, as explained above, it is essential to assess the possibility that symbols can unduly compensate for unfavorable or logically unpersuasive information that should yield negative responses.

@@

According to ELM, the responses of individuals to persuasive communications depends on the amount of thinking that they dedicate to the message content. Persuasion can occur when thinking is "high" or "low", but the processes and consequences are likely to differ in each situation. This range between high and low degrees of thought is referred to as the "elaboration continuum". People's location along this continuum is associated with different psychological processes of persuasion. At the high end of the elaboration continuum, also called the "central route", people are likely to respond by cognitively scrutinizing the arguments central to the merits of the issue. At the low end, also called "the peripheral route" people respond to simple cues automatically and unconsciously through primitive psychological processes that require very little amount of cognitive effort, such as an unconscious transfer of affect[^footnote_dual_process_theories]. In turn, these different persuasion mechanisms are likely to cause divergent types of attitudinal change. Persuasion through the central route is expected to be more enduring, resistant and predictive of behavior, compared with peripheral route persuasion. Of course, much of the time, people are located in between the two ends of the continuum, and thus persuasion is determined by a mixture of these processes.

@@

ELM enables to disentangle the distinct effects of different elements in communications. Some elements are processed differently by people at different locations along the elaboration continuum. Consider for instance a text that contains a series of logical arguments supporting a certain position. Those at the high end of the continuum would regard this text as arguments, and systematically scrutinize it based on its logical persuasive quality (central route processing). Conversely, those at the low end of the continuum are likely to disregard the content of the text, and instead, rely on the number of lines in the text, for example, as a cue for its strength (peripheral route processing). Some elements provide potential cues that lack any logical quality at all. For example, the physical attractiveness of the source normally does not represent any substantive argument, yet it can serve as a simple cue. These latter elements may affect the attitudes of those at the low end, but are less likely to persuade those at the high end [@petty_1986].

ELM further postulates that the placement of people along the elaboration continuum in any given situation is determined by their *motivation* and *ability* to think about the message. Accordingly, the literature proposes a variety of such motivational and ability factors. Among others, studies have demonstrated that the more people are distracted, the more difficult it is for them to scrutinize the message arguments, a mechanism that has been referred to as "elaboration disruption" [@petty_1976; @petty_1986]. Another prominent factor that influences people's inclination to elaborate on the message is their affective state, or mood. Positive mood (compared with neutral or negative moods) reduces people's cognitive ability and attenuates their motivation to think, due to their desire to maintain their positive mood and avoid message processing that would depress it [@worth_1987; @mackie_1989; @bless_1990; @schwarz_1991; @petty_2015; but see: @wegener_1995].[^footnote_elm_factors] Affecting the amount of thinking can cause attitude change in interaction with elements in the communication that can function as arguments or peripheral cues. Decreasing the amount of thinking weakens the effect the substantive arguments and enhances the effect of peripheral cues (and vice versa).

@@

Elements in communications can also function as factors that affect people's motivation and/or ability to think. For example, Background music in commercials can distract people and prevent them from scrutinizing their content [@park_1986]; The use of humor in television shows stimulates positive mood and reduces people's ability and motivation to scrutinize their substantive arguments [@nabi_2007; @young_2008]. Hence, elements can have more than one function, and can shape attitudes through several mechanisms. They can serve both as variables that affect the amount of thinking and as potential peripheral cues or arguments. Continuing with the above-mentioned examples, background music and humor can serve as simple cues that persudate people directly through the peripheral route, yet at the same time they can also as factors that reduce elaboration. Accordingly, they can affect people both directly (as peripheral cues) and indirectly (by decreasing elaboration and thus attenuating the effect of arguments and enhancing the effect of peripheral cues).[^footnote_multifunction_factors]

[^footnote_multifunction_factors] : @petty_1986 have suggested that we are more likely to find complex relations of multiple mechanisms in those situations where people are located in the middle of the elaboration continuum. When people are neither extremely motivated or unmotivated to think, they may rely on different elements in the communications to decide if the message is worth or needs considering.

@@

In light of the conceptualization of ELM, substantive information elements in government public communications represent issue-related arguments, which can be scrutinized or processed systematically based on their logical quality (via the central route), at least by those who are sufficiently motivated and capable to do so. Conversely, symbolic elements are potential peripheral cues, which can affect citizens' trust through the peripheral route, by evoking positive emotions and associations that are attributed to government organizations and their policies [@alonbarkat_2017]. However, in addition to their direct effect as peripheral cues, symbolic elements may also increase citizens' trust indirectly. They can decrease the amount of thought citizens dedicate to the communication, and thus prevent them from scrutinizing negative information (e.g. information on poor performance) and weak arguments (e.g. logically unpersuasive description of their policy plans).

Symbolic elements may decrease citizens' ability and motivation to think about the message in several ways. First, they can cause elaboration disruption, by distracting citizens and absorbing their attention and cognitive efforts [@petty_1976; @petty_1986]. Second, they can decrease citizens' motivation and ability to think about the message content by creating a positive affect and placing citizens in a positive mood. As explained above, the postulation that positive mood decreases elaboration has been explained, in part, by a mood maintenance mechanism [@mackie_1989; @bless_1990; @schwarz_1991]. This mechanism is particularly relevant to the context of government communications, in which thinking about the content of the message - government organizations and policies - is likely to yield negative mood among most people [@marvel_2015; @marvel_2015a; @hvidman_2016]. Therefore, when citizens are exposed to symbolic elements that evoke positive feelings, they may seek to avoid thinking about tedious bureaucratic affairs, which would depress their positive mood. The latter expectation entails an interaction between symbols and the substantive information in the communications. Symbolic elements are likely to increase citizens' trust insofar as they reduce citizens' inclination to scrutinize information that would have decreased their trust.

In sum, I expect that:

@@

>*H<sub>2</sub> - The positive effect of symbolic elements on citizens' trust is likely to be enhanced when the communication includes negative or logically unpersuasive information that would have decreased citizens' trust.*

>*H<sub>3</sub> - Symbolic elements are likely to decrease the amount of thinking about the content of the communication.*

@@

Analytically, the expectation that symbolic elements decrease message elaboration also entails that symbols may decrease citizens' trust in communications about highly positive performance and persuasive policy plans. Nevertheless, one can make the argument that neutralizing the effect of "positive" information with symbols may not actually yield significant differences, since citizens are more likely to "punish" bureaucracies for publishing negative information than to "praise" them for publishing positive information [@james_2011; @nielsen_2016; @james_2017b]. Congruently, the effect of positive information (versus a medium level) is likely to be relatively small, and hence neutralizing this effect may not make much difference. Given the focus of this study on the interaction between symbols and ill-conceived policies, the research design does not directly addresses this latter interaction.

@@

In addition to the abovementioned hypotheses, the experiment is also designed to test an additional set of hypotheses regarding the interaction between symbols and the perceived personal relevance of communications. The selection of policy plans and the construction of the population sample were designed to account for variation in that variable. In this paper, I focus on those methodological and empirical parts of the study that are consequential for the examination of the interaction bewteen symbols and information.[^footnote_relevance_paper]

[^footnote_relevance_paper]: The methodological aspects which are relevant for examining the additional hypotheses (i.e. the interaction between symbols with personal relevance) are described in a seperate paper. An English translation of the full survey is available from the supplementary appendix.

@@

Empirically, I focused on the case of the Israeli Environment Protection Ministry (hereafter: EPM), a national level bureau responsible for the formulation, co-ordination and execution of environmental policy in Israel, both at the national and local levels. I analyzed the effects of informative and symbolic elements in EPM's communications of two policy plans on citizens' trust in these policies, and the extent to which they elaborated on them. In the survey, the communications about the two policy plans, in which I manipulated the informative and symbolic elements, are presented to the participants as sections from the EPM's official publication of its annual work plan.

Methodologically, EPM makes a good case for this study since has invested greatly over the past years in public communications. In these various communications, the EPM used an array of symbolic elements, which became widely recognized by the public. Here, I focused on three prominent symbolic elements used in the EPM's public communications, and displayed below in **Figure 2**. The first symbolic element is the unique brand logo of the EPM (image a. in figure 1). This logo consists of pair of green and orange leaves that resemble two hands, designed to symbolize peace and harmony with the environment. The second symbolic element in which I focused is the green color, which is strongly associated with taking care of the environment. The EPM's logo and the green color can be seen in almost all of the visual communications of the EPM, and they represent key elements of its unique "visual identity" [@vandenbosch_2005]. The third element is celebrity endorsements. Specifically, I selected two high-profile celebrity comedians, *Tal Friedman* and *Ido Rosenblum*, that previously endorsed two salient advertising campaigns of the EPM. The campaign starring the comedian Tal Friedman (image b.) was launched in 2010, and included a series of advertisements in which Friedman was presented with a green grass head, and accordingly the campaign's slogan was: "starting to think green". The campaign starring the comedian Ido Rosenblum (image c.) was launched in 2017, and was focused on reducing the use of disposable bags.[^footnote_campaigns] The two campaigns were highly successful in terms of their recognizability and likability, as suggested by their evaluation reports by the Government advertising agency.

@@

## *Procedure and sample population*

The survey was presented to the participants as a study about citizens' attitudes regarding environmental policy issues. The participants were first asked pre-manipulation questions about their political ideology and their perceptions about the pubic sector.[^footnote_relevance_manipulation] Thereafter, they were told that they will be presented with two government policy plans regarding specific environmental issues, obtained from EPM's official publication of its annual work plan for the year 2018. The policy plans regarded to reducing air-pollution in the Bay of Haifa (an area in Israel) and reducing domestic waste and increasing recycling. The two policy plans were presented to each subject in a random order. Each policy plan included the title of the policy goal (i.e. "reducing the levels of air pollution in the Haifa bay"), followed by short descriptions of two actions to be taken by EPM in order to fulfill this goal (30-50 words each).

[^footnote_relevance_manipulation]: This section also included a manipulation of participants' perceived personal relevance. Participants in the treatment group were asked a series of questions about environmental governent policy, whereas participants in the control group were asked instead equivalent questions about their field of occupation.

Per each policy, the subjects were randomly allocated to one of two conditions of substantive information (hereafter: information): strong policy (i.e. logically persuasive policy plan) or weak policy (i.e. logically unpersuasive). The policy activities which were mentioned in the strong policy plans condition were taken from the real EPM work plan and from other reports and official sources of EPM. The activities that were included in the weak policy plans condition were fictional, and were designed to represent measures which are likely to be perceived by subjects as extremely incompatible with achieving the policy goals, yet still sufficiently reliable. For instance, the proposed actions for the air-pollution strong policy involved increasing the supervision of polluting factories, and reducing diesel smoke emissions by vehicles, whereas the actions for the weak policy were increasing the supervision of polluting factories, and raising citizens' awareness of the Ministry's efforts to improve air quality. The full texts of the weak and strong conditions for the two policy plans is displayed in **Appendix A**. For each subject, one of the two policies was strong and the other was weak. For example, a subject who first received a strong air-pollution policy, later received the weak recycling policy, and vice versa.

In addition to information manipulation, the participants are randomly assigned to one of three conditions of symbolic elements in the communication (hereafter: symbols): treatment (real symbols), control no symbols and control fake symbols. In the symbol treatment condition, the two policies (again, one weak and one strong) are displayed with the above-mentioned three symbolic elements: the titles and the footer are colored with green, EPM logo is displayed at the top of the page, and the celebrities' images from the campaigns are attached and labeled "EPM advertisement". Each communication included the image of one celebrity: Rosenblum's image from the campaign for reducing the use of disposable bags is attached to the recycling policy; Friedman's image from the "starting to think green" campaign was attached to the air-pollution policy. In the control no-symbols condition, the two policies are displayed in a minimal black and white design, and without the logo and the celebrities images. Finally, in the control fake-symbols condition, the green color is replaced with blue, the EPM logo is replaced by a fake logo and the images of celebrities from the campaigns are replaced with two edited images of unfamiliar people. The purpose of the fake-logo condition is to isolate the symbolic meaning of the EPM's graphic identifiers and distinguish it from the possible effect of their aesthetic design. Accordingly, I specifically designed the "fake" symbols to that they will resemble the real symbols and have equivalent aesthetic qualities, without activating emotions and associations which can be attached to the government ministry and environmental issues.[^footnote_symbols_manipulation_checks] In that sense, it can be regarded as a placebo. The symbols conditions are displayed in **Appendix B**. That each subject received two policies which were presented under the same symbol condition, but under different information conditions (i.e. one strong and one weak), enables me later to analyze the interaction between these two manipulations both via between- and within-subjects.

[^footnote_symbols_manipulation_checks]: In two prior surveys, I empirically confirmed that the fake symbols are unfamiliar, do not have strong positive or negative affect, and are not strongly associated with environment policy and EPM. The results of these examinations are reported in the supplementary appendix (##TBA##).
